---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# historicalborrow

[![check](https://github.com/wlandau/historicalborrow/workflows/check/badge.svg)](https://github.com/wlandau/historicalborrow/actions?query=workflow%3Acheck)
[![cover](https://github.com/wlandau/historicalborrow/workflows/test-coverage/badge.svg)](https://github.com/wlandau/historicalborrow/actions?query=workflow%3Atest-coverage)
[![pkgdown](https://github.com/wlandau/historicalborrow/workflows/pkgdown/badge.svg)](https://github.com/wlandau/historicalborrow/actions?query=workflow%3Apkgdown)
[![lint](https://github.com/wlandau/historicalborrow/workflows/pkgdown/badge.svg)](https://github.com/wlandau/historicalborrow/actions?query=workflow%3Alint)

Historical borrowing in clinical trials can improve
  precision and operating characteristics. This package supports
  a hierarchical model and a mixture model to borrow historical
  control data from other studies to better characterize the
  response of the control arm of the current study. It also quantifies
  the amount of borrowing through benchmark models (independent
  and pooled). Some of the methods are borrowed from
  [Viele et al. (2013)](https://onlinelibrary.wiley.com/doi/10.1002/pst.1589).

## Installation

```{r, eval = FALSE}
remotes::install_github("wlandau/historicalborrow")
```

## Documentation

* Functions: <https://wlandau.github.io/historicalborrow/reference/>
* Methods: <https://wlandau.github.io/historicalborrow/articles/methods.html>
* Usage: <https://wlandau.github.io/historicalborrow/articles/usage.html>

## Thanks

[Albert Man](https://github.com/albert-man) reviewed this package. Albert, [Faith Bian](https://github.com/faithbian-lilly), and [Saptarshi Chatterjee](https://github.com/schatterjee-lilly) contributed to the development of the methods prior to implementation. Phebe Kemmer, Heather Zhao, and Zhangchen Zhao also provided helpful feedback on the models and their application to clinical use-cases.

## Code of Conduct

Please note that the historicalborrow project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
